HO unification from object language to meta language

### 2024-03-30 paper/main.tex
rework maybeeta example

Compiling a term $t$ forces us to determine if there exists a $t' \in \subterm{t}$,
such that $t' = \lambda x. t''_x$,
for any term $t''$, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.t''_x) = t'$. This
\maybeeta detection is not a trivial operation as it may seems.
\\
\noindent
\begin{minipage}{.5\linewidth}%
  \begin{gather}
    \lambda x. f \appsep A_x\\
    \lambda x. f \appsep x \appsep A_x
  \end{gather}
\end{minipage}% This must go next to `\end{minipage}`
\begin{minipage}{.5\linewidth}
  \begin{gather}
    \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
    \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
  \end{gather}
\end{minipage}

\begin{center}
  \begin{tabular}{lll}
    Term & Status & Evidence \\\hline
    $\lambda x. f \appsep A_x$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto x ~\}$ \\
    $\lambda x. f \appsep x \appsep A_x$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep A_x \appsep B_{y x}$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto x,~ B_{yx} \mapsto y ~\}$\\
    $\lambda x. f \appsep A_x \appsep x$ & $\in\maybeeta$ & $\sigma = \{~ A_x \mapsto a ~\}$
  \end{tabular}
\end{center}

In the examples above, the first term is a \maybeeta since $A_x$ can
reduce to $x$ by setting $A_x = \lambda x.x$, 
the second one is not a \maybeeta since it exists no substitution
for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
It is a bit more complicated to determine if eq.~\ref{eq:lamlam} a \maybeeta since,
we have a spine of lambdas, this means that the whole term
is a \maybeeta, if the inner $\lambda$-term $\lambda y.f\appsep A_x\appsep B_yx$,
is an $\eta$-expansion and, if so, we need to verify if the $\eta$-reduced 
subterm, i.e. $f \appsep A_x$, can make $\lambda x.f \appsep A_x$ an $\eta$-expansion.
Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.
Lastly, eq.~\ref{eq:lamdisc} is a \maybeeta even if $x$
occurs in the subterm $t' = f \appsep A_x$, since it exists a substitution
$\sigma = \{A \mapsto \lambda x.a\}$, for any constant $a$, such that $\sigma t'
= f \appsep a$ is a term where $x$ does not appears as a free variable.

>>>>

When compiling a term $t$ we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems, here a few examples:
% \\
% \noindent
% \begin{minipage}{.5\linewidth}%
%   \begin{gather}
%     \lambda x. f \appsep A_x\\
%     \lambda x. f \appsep x \appsep A_x
%   \end{gather}
% \end{minipage}% This must go next to `\end{minipage}`
% \begin{minipage}{.5\linewidth}
%   \begin{gather}
%     \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
%     \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
%   \end{gather}
% \end{minipage}
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
The third example shows when a variable occurs outside the scope of a unification
variable cannot be erased and hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself, that means it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\
